Density-dependent selection and the limits of relative fitness

### 2018-02-24 manuscript.tex
l

\section*{Appendix B: Derivation of growth equation}

@@

When making approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, we must ensure that the resulting growth equations are logically consistent. For instance, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories make a negligible contribution to growth. Consequently, the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. The approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must then be exactly equal for logical consistency. Even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually excellent approximations, without this consistency condition the resulting approximate model implies that the rare type grows or declines purely by virtue of being rare, but we could select any rare subset of a single-type population and predict the same behavior --- a contradiction.

To be more concrete, suppose that we naively use $\tilde{p}$ and $\hat{p}$ instead of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively. Then, for the rare type, the mean total propagule density in territories were it is present is $\langle x_2\rangle_{\tilde{p}}+1\approx L+1$ (since the rare and common types disperse independently), and the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$.

in the Eq. \eqref{eq:meanfieldr} approximation.

On the other hand, virtually all nonfocal propagules

$\tilde{q}$ and $\hat{q}$ are chosen to

$\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation

@@

Since $x_i$ and the $x_j$ are independent under $\tilde{p}$

If the focal type is rare, the mean propagule density is $L+1$ in territories with one focal propagule.

@@

$\tilde{p}$

The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal. $\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).

>>>>

\section*{Appendix B: Growth equation derivation}

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. Suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories make a negligible contribution each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, these identical types should retain constant relative frequencies, and so the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.

This consistency requirement is not automatically satisfied even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually good approximations. Even small violations will cause one of the identical types to grow exponentially relative to the other. This pathological growth occurs purely because one group is treated as rare and the other as common. Since a single-type population can be divided into arbitrary rare and common subsets of identical types, and clearly not all such subsets can grow/decline, we have a logical contradiction.

To be more concrete, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively. Then, since $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type. And, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Therefore, the consistency requirement is violated if we naively use the conditional distributions $\tilde{p}$ and $\hat{p}$ derived from the dispersal distribution $p$ to calculate the effective means.

The effective distributions $\tilde{q}$ and $\hat{q}$ are chosen to ensure that this consistency requirement is satisfied.

$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).

@@

@@
